---
title: "Gapminder-Exploration"
output: 
  html_document:
    keep_md: yes
    theme: cerulean
---
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(gapminder))
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r cars}
library(gapminder)
library(tidyverse)
```

# Tasks

## Task 1 

The first thing I will do is extract the minimum and maximum GDP per capita for each of the continents. I then extract the minimum and maximum life expectancies across all the continents. I will use the group by and summarise functions to extract the necessary data.  I then plot both the min and max life expectancies for all continents.
```{r}
df <- gapminder %>%
  group_by(continent) %>%
  summarize(min = min(gdpPercap),max = max(gdpPercap)) 
df # print table

df2 <- df %>%
  summarize(minAll = min(min),maxAll = max(max))
df2 # print table

ggplot(df, aes(continent,min)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Minimum Life Expectancy for each Continent") + 
  ylab("Min Life Expectancy") # plots min life expectancy for each continent
ggplot(df, aes(continent,max)) + 
  geom_bar(stat = "identity") + 
  ggtitle("Maximum Life Expectancy for each Continent") + 
  ylab("Max Life Expectancy") # plots min life expectancy for each continent
```

## Task 2

I now compare the spread of GDP per capita across the various continents, by computing the maximum and minimum GDP per capita and taking the difference. I used a boxplot since it allows you to compare spread of data very easily. 
```{r}
df <- gapminder %>%
  group_by(continent) %>%
  summarize(diff = max(gdpPercap) - min(gdpPercap)) # compute difference between max and min GDP per capita for each continent
df # show table
ggplot(gapminder, aes(continent, gdpPercap)) + 
  geom_boxplot() + ggtitle("Comparison of GDP per capita by Continent") + 
  ylab("GDP per capita") 
```

As we can see from both the boxplots and the table, Asia has by far the largest spread of GDP per capita's.

Here we examine the number of countries in each continent that have a life expectancy over 50, and how this changes over time.

## Task 3

Here I compare how mean life expectancy for each continent is changing over time. I chose to use the average life expectancy to avoid having the plots become too cluttered. As well, I chose to use scatterplots, as I am plotting 2 quantitative variables.
```{r}
df <- gapminder %>%
  group_by(year,continent) %>%
  summarize(meanlifeExp = mean(lifeExp)) # group by continent and year, and compute mean life expectnacy for each grouping
df # print table
ggplot(df, aes(year, meanlifeExp)) + 
  geom_point() + 
  facet_wrap(~ continent) + 
  ggtitle("Mean Life Expectancy versus Year for each Continent") +
  ylab("Mean Life Expectancy") # plot mean life expectancy versus year for each continent
```

Overall, it seems like all continents have gradually increasing life expectancies. However, Africa seems to have reasonably leveled off in roughly the last 20 years.

## Task 4

```{r}
df <- gapminder %>%
  group_by(year,continent) %>%
  filter(lifeExp > 50)
df <- count(df) # stores number of countries with a life expectancy > 50 in a given continent, for each year, in a variable called n
df # prints table
ggplot(df, aes(year, n)) + 
  geom_point() + 
  facet_wrap(~ continent) + 
  ggtitle("Count of countries in each year with life expectancy over 50") + 
  ylab("Count") # rename n to count

```

As we can see, Africa and Asia have the largest increase over time in terms of the number of countries having life expectancies over 50. The Americas have a smaller change, and Europe and Oceania have almost no change.

## Task 5

I chose to attempt the open-ended task. I took the difference between lag life expectancy and life expectancy, and sort countries in descending order, in order to determine which countries had the largest drops in life expectancy (over 5 year span). I only plotted those countries where the life expectancy dropped by more than 6, to keep the plot from becoming too Scluttered.
```{r}
df <- gapminder %>% # extracts and sorts largest drops in life expectancy by country
  group_by(country) %>%
  mutate(change = lag(lifeExp) - lifeExp) %>%
  filter(!is.na(change), change > 6) %>%
  summarize(popIncrease = max(change)) %>% 
  arrange(desc(popIncrease))
df # print out table
ggplot(df, aes(country,popIncrease)) +  
  geom_bar(stat = "identity") + 
  ggtitle("Max Drop in Population over 5 Year Span") + 
  ylab("Population Change") # to create bar chart comparing country life expectancy
```

As we can see, Rwanda had the largest drop in life expectancy across all countries. This is likely due the civil war in the country in the early 90's. Various other countries on this list had other horrific events happen as well.